Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-03-07 thesis/example.tex
not working!!

\lhead{Ningrui Xie}
\chead{BIO 4C12 Midyear Report}
\rhead{ID: 400323416}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0pt}

\newcommand{\perday}{\ensuremath{/mathrm{day}}

@@

\section{Results}
In Figure 1, we present the comparison plots between the infectious stage duration distribution generated by the SI$^n$R model with the Erlang distribution. In this model, the mean recovery time is set to be 10 days $(\gamma = \frac{1}{10})$ and the infectious stage is divided into $n$ substages $(n = 4, 8, 12, 18)$, with each substage having a rate of $\frac{n}{10}$. The corresponding shape and scale parameters for the gamma distribution are $n$ and $\frac{n}{10\perday}$, respectively. The plot demonstrates a perfect match between the two distributions, despite being generated by different methods. In the subsequent sections, we will refer to these two distributions interchangeably.
\begin{figure}[h]
    \centering
    \includegraphics[width= 0.9\textwidth]{4.png}
    \caption{Probability Distribution Comparison Between Erlang and SI$^n$R Model}
\end{figure}
JD: Respect the units!

\subsection{Possible match between SI$^n$R and SIgR}
In the initial step following the successful construction of the SIgR model, which employs a fixed number of substages and a geometrically distributed substage rate, we want to evaluate the theoretical possibility of a match between the stage duration distribution produced by the SIgR model and that of the traditional SInR model. This match is defined in terms of two key parameters: the mean $(M)$ and the shape parameter $(\kappa)$.

The substage number of the SIgR model $(n_f)$ has been fixed at 12, and the mean stage duration for the entire infectious period is set to 10 days. We examine the effect of using 2, 4, and 8 infectious substages in the SI$^n$R model $(n = 2, 4, 8)$ on the values of $r$ required for the SIgR model to align formulaically with the shape parameter $(\kappa)$. In Figure 2, the plots on the left illustrate the changes in stage duration distribution in SIgR model resulting from varying $r$ within a range that includes the formulaic match for $\kappa$. The black curve represent duration distribution from the SI$^n$R model with the associated $n$. Plots on the right depict the variations in $\kappa$, calculated using the corresponding r values, where the red dashed line is the target value of $\kappa$. It is evident from the plots that, within the specified range of r, there is a concurrence in both the mean and shape parameter values between the SInR and SIgR model. These findings suggest that the results generated by the traditional mode are replicable using the model we propose.

\begin{figure}[h!]
    \centering

    \begin{subfigure}{0.6\linewidth}
       \includegraphics[width=\linewidth]{4.1(n=2).png}
    \end{subfigure}

    \begin{subfigure}{0.6\linewidth}
       \includegraphics[width=\linewidth]{4.1(n=4).png}
    \end{subfigure}

    \begin{subfigure}{0.6\linewidth}
        \includegraphics[width=\linewidth]{4.1(n=8).png}
     \end{subfigure}

    \caption{Adjusting $r$ to Evaluate Match While Preserving Mean Value}
    \label{fig:all_plots}
 \end{figure}

\subsection{Simple Equations Facilitating Independent and Flexible Changes in Mean and $\kappa$}
In the SIgR model, the calculation of the Mean $(M)$ and Variance $(\sigma^2)$ for the infectious stage duration distribution is accomplished by adding the corresponding properties of each substage, as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{M}$, can thus be simplified to:
\begin{align}
    \kappa = \frac{(1+r^{n_f}) (1-r)}{(1+r) (1-r^{n_f})}
\end{align}
This simplification reveals that the value of $\kappa$ is exclusively dependent on the parameter $r$, thereby granting the model independence in property changes. This independence allows for the alignment of $\kappa$ without altering the Mean, achievable through adjustments to parameter $a$. Similarly, the Mean can be scaled to a desired value without impacting $\kappa$.

JD: I would frame this differently. Both PE and Erlang \emph{can} change the value of $n$; the real difference is that PE can be flexible even when $n$ doesn't change.
Importantly, it should be noted that the value of $\kappa$ is not affected by changes in the number of substages, as $n_f$ is fixed. Consequently, $\kappa$ can be any real number within the range $[\frac{1}{n_f}, 1)$, thereby significantly enhancing the model's flexibility. In contrast, in the SI$^n$R model, the value of $\kappa$ depends on how the substages are partitioned. Following the same steps of calculation, we obtain: $\kappa = \frac{1}{n}$. The shape parameter $\kappa$ in the SI$^n$R model is therefore restricted to a set of discrete values:
\begin{align*}
    \kappa \in \left\{1, \frac{1}{2}, \frac{1}{3}, \dots \right\}
\end{align*}
This limitation potentially reduces the capability of the SI$^n$R model to accurately reflect real-world scenarios.

\subsection{Identifying Parameters $a$ and $r$ to Align SI$^n$R and SIgR with Given Mean and $\kappa$}
Utilizing formulas (4) and (6), we have developed a computational system to reverse-engineer the process. The parameters $a$ and $r$, generated by this system, were reintegrated into the SIgR model. This enable us to compare the stage duration distribution with that of the SI$^n$R model, which possesses identical mean and shape parameters, as depicted in Figure 3. Although the formulaic $M$ and $\kappa$ are computed to be the same, we also conducted a numerical calculation of both properties, with results presented in Table 1. This analysis confirms that a match is achieved at both the formulaic and numerical levels. The visual discrepancies observed between the two distributions in Figure 3 can be considered negligible, since both distributions serve as theoretical representations of real-world scenarios and do not require exact correspondence.
\begin{figure}[h]
    \centering
    \includegraphics[width= \textwidth]{4.3.1.png}
    \caption{Theoretical Alignment of Mean and Shape Parameters in SI$^n$R and SIgR Models with Calculated Parameter Values $a$ and $r$}
\end{figure}

\begin{table}[h!]
    \centering
    \begin{tabular}{ l | c | r }
      \hline
      Substages & Mean & $\kappa$ \\
      \hline
      n = 2 & 10.000000 & 0.4999971 \\
      n = 4 & 10.000000 & 0.2500002 \\
      n = 8 & 10.000 & 0.125 \\
      n = 12 & 10.000000 & 0.0999986 \\
      \hline
    \end{tabular}
    \caption{Numerically Calculated Parameter Values From the SIgR Model}
\end{table}

JD: A table caption should have a clear explanation of what's in the table for people who are skimming (your figure captions should probably be improved, too). It's not clear what you're showing here, or whether it's right. Maybe show target values as well as observed values? Also, there seems to be a mismatch between $n=12$ and the corresponding $\kappa$? Also, there are too many significant figures here. We don't need to fix this table, since we don't need it moving forward, these are just general tips.

By incorporating the infectious stage duration distribution back into the entire SI$^n$R and SIgR models, we also present the dynamics of the infectious population density generated directly from the simulations. In Figure 4, using $n=2$ and $n=4$ for the SI$^n$R model and corresponding $a$ and $r$ values ( $r_2=2.9999604, a_2=0.1500007, r_4 = 1.6626998, a_4 = 0.2503359$) for the SIgR model, a close match is observed between the results generated by the two different methods. This indicates that the results of the SInR model are reproducible using the SIgR model, but with enhanced flexibility in parameter value changes.
\begin{figure}[h]
    \centering
    \includegraphics[width= \textwidth]{4.3.2.png}
    \caption{Simulation Comparison of the Population Dynamics Between SI$^n$R and SIgR Model with Equivalent Mean and Shape Parameter}
\end{figure}

\subsection{Enhanced Flexibility in Controlling Stage Duration Distribution Shapes}
Besides offering a simpler approach to constructing an epidemic model with efficacy comparable to the SInR model, the SIgR model further enhances our ability to manipulate the shape of the stage duration distribution. This is achieved by allowing the selection of any $\kappa$ value within the range $(\frac{1}{n_f}, 1)$. Figure 5 showcases the probability distribution plots generated by the SIgR model, given a desired mean $(M=10)$ and selected $\kappa$ value $(\kappa = 0.25, 0.45, 0.65, 0.85)$. This chosen $\kappa$ value (except 0.25) is unattainable in the SI$^n$R model, distinguishing the SIgR model's unique capabilities. The new model adeptly incorporates a wider array of distribution shapes into the SIR framework. It emerges as a versatile tool for capturing a more extensive range of dynamic patterns in disease spread and control.
\begin{figure}[h]
    \centering
    \includegraphics[width= \textwidth]{4.4.2.png}
    \caption{Generating Stage Duration Distributions with Custom Mean and Shape Parameters Using the SIgR Model}
\end{figure}

\section{Discussion}
SIR compartmental models are widely utilized in modeling infectious diseases, but they encounter a significant limitation: the implicit assumption that the time an individual spends in the disease stage follows an exponential distribution. To address this issue, a common method involves subdividing the infectious stage into $n$ number of substages, each substage is exponentially distributed with a constant rate. This approach is also known as the Linear Chain Trick (LCT). LCT strategy facilitates an Erlang-shaped stage duration distribution with increased reality. However, this model also faces limitations, including cumbersome parameter optimization processes and the constraint of discrete parameter values, which can affect its performance.

In this study, we propose the SIgR model, which features a fixed number of infectious substages and a geometric distribution rate for these substages, as an alternative to the SInR model that incorporates LCT. In the results section above, we demonstrated that adjusting the parameters for the geometric series $a$ and $r$ allows us to replicate the simulation results generated by the SInR model. Moreover, our model offers enhanced flexibility in shaping the distribution of the infectious stage duration, as it allows for the selection of mean $(M)$ and shape parameters $(\kappa)$ of the distribution such that $M \in \mathbb{R}^+$ and $\kappa \in [\frac{1}{n}, 1)$. Additionally, the fixed structure of our model simplifies the process of finding optimized parameter values, requiring only a one-time process during data fitting. This simplification streamlines the development of computer code for use with ODE solvers in R, ultimately saving time and reducing complexity.

Future work will primarily focus on fitting data to the model we have proposed. This will include both synthetic data generated through simulation and actual real-world data. The results of this fitting process will be compared with those obtained from the SInR model. Later on, we plan to incorporate process error considerations to increase the robustness of our model.

The anticipated outcome of this research is the development of a model that not only exceeds current models in terms of predictive accuracy but also maintains simplicity in implementation. The core design of our model is intended to be easily adaptable across various state transition models. This adaptability is vital for its application in diverse domains, ranging from epidemiology to system dynamics. The idea is designed to be extensible, capable of accommodating a range of distribution types and parameter sets. Such versatility makes it a powerful tool for modeling a wide array of complex systems, with the capability to adjust and evolve in response to new data and insights.

In conclusion, this study suggests that our method could simplify and enhance the fitting of models with flexible time distributions within an Ordinary Differential Equation (ODE) framework. This research has the potential to enhance model performance and to develop user-friendly tools for effective epidemic modeling and prediction.

\bibliographystyle{unsrt}
\bibliography{Reference}

>>>>

\lhead{Ningrui Xie}
\chead{BIO 4C12 Example document}
\rhead{ID: 400323416}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0pt}

\newcommand{\perday}{\ensuremath{/mathrm{day}}}

@@

\section{Results}
In Figure 1, we present the comparison plots between the infectious stage duration distribution generated by the SI$^n$R model with the Erlang distribution. In this model, the mean recovery time is set to be 10 days $(\gamma = \frac{1}{10})$ and the infectious stage is divided into $n$ substages $(n = 4, 8, 12, 18)$, with each substage having a rate of $\frac{n}{10}$. The corresponding shape and scale parameters for the gamma distribution are $n$ and $\frac{n}{10\perday}$, respectively. The plot demonstrates a perfect match between the two distributions, despite being generated by different methods. In the subsequent sections, we will refer to these two distributions interchangeably.
\begin{figure}[h]
    \centering
    \includegraphics{figures/test.Rout.pdf}
    \caption{Probability Distribution Comparison Between Erlang and SI$^n$R Model}
\end{figure}

\bibliographystyle{unsrt}
\bibliography{Reference.bib}
